---
title: "Missing Data Imputation"
author: "Hun"
date: '2022-04-28'
output: pdf_document
---


```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE
)
```

```{r}
library(tidyverse)
library(mice)
library(visdat)
library(VIM)
library(patchwork)
```

```{r}
x <- RCurl::getURL("https://raw.githubusercontent.com/TarekDib03/Analytics/master/Week3%20-%20Logistic%20Regression/Data/framingham.csv")

data <- read.csv(text = x) %>% janitor::clean_names()
```


## Visualizing missing data

```{r}
# Visualizing missing data
vis_miss(data)
```

## Patterns of missing data

```{r}
aggr_plot <- aggr(data, col=c('skyblue','orange'), numbers=TRUE, sortVars=TRUE, labels=names(data), 
                  cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))

```

\newpage  
### Checking the assumption of missing completely at random (MCAR)

If our assumption of MCAR data is correct, then we expect the red and blue box plots to be very similar.

```{r, fig.height=4, fig.width=5}
# If our assumption of MCAR data is correct, then we expect the red and blue box plots to be very similar.
marginplot(data[c(15,2)])
```


```{r, fig.height=4, fig.width=5}

marginplot(data[c(15,11)])
```


### Selecting variables to impute data

I selected *glucose* and *tot_chol* because they both are continuous and have similar patterns.

```{r}
selected_varaible <- 
  data %>% 
  dplyr::select(-education, -bp_meds, -cigs_per_day, -bmi, -heart_rate)
```

### Using Predictive Mean Matching (PMM) to impute missing data

```{r}
tempData <- mice(selected_varaible, m = 5, maxit = 5, meth = 'pmm',seed = 7)

```


```{r}
completedData <- complete(tempData,1)
```

```{r, fig.height=4, fig.width=8}
xyplot(tempData, glucose ~ sys_bp + age + dia_bp,pch=18,cex=1, 
       main = "Red: imputed glucose vs. Blue: observed glucose")
```


```{r, fig.height=4, fig.width=8}
xyplot(tempData, tot_chol ~ sys_bp + age + dia_bp,pch=18,cex=1, 
       main = "Red: imputed tot_chol vs. Blue: observed tot_chol")
```

```{r}
densityplot(tempData, main =  "Red: the imputed vs. Blue: the observed")
```

